Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comprehensive Exercise: Carleton College}

### 2010-03-07 Paper.tex
Fixed code and whitespace problems.

% This should be a lot prettier.
\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota \\
{\large Senior Comprehensive Exercise: Carleton College}}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

@@

    \begin{figure}
    \begin{center}
        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
        \caption{A visualization of how the hierarchical model works.}
        \label{simplehierarchicalmodel}
    \end{center}
    \end{figure}

@@

    \begin{figure}
        \centering
        \includegraphics[width=5in]{observerPlots.pdf}
        \caption{The box plots show how the proportion of plots in which an
        observer saw tracks (``occupancy'') differs among the three observers in
        2003. Only those dates on which all three observers flew are included.}
        \label{obsPlots}
    \end{figure}

@@

    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "irregularHierarchicalModel.txt")
modelCheck("irregularHierarchicalModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal"), 
         "irregularOtterData.txt")
modelData("irregularOtterData.txt")
modelCompile(numChains=2)
modelGenInits()

@@

\section{CAR model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    for (site in 1:sites) {
        logit(theta[site]) <- alpha1[site] + alpha0
    }
    for (i in 1:sumNumNeigh) {
        weights[i] <- 1
    }
    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
    alpha0 ~ dflat()
    tau ~ dgamma(.5,.0005)
    psi ~ dunif(0,1) 
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    #c ~ dgamma(1,.1)
    #d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta(a,b) %_% I(.6, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .3)
    }

@@

    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "irregularHierarchicalModel.txt")
modelCheck("irregularHierarchicalModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal", "num", "adj",
           "sumNumNeigh"), 
         "irregularOtterData.txt")
modelData("irregularOtterData.txt")
modelCompile(numChains=2)

init.val <- function(){
	list(
		alpha1=runif(sites,-100,100),
		alpha0=runif(1,-100,100),
		tau=rgamma(1,.5,.0005),
		psi=runif(1),
		a=rgamma(1,1,.1),
		b=rgamma(1,1,.1)
		#,c=rgamma(1,1,.1),
		#d=rgamma(1,1,.1)
	) 
}
bugsInits(init.val,numChains=2,fileName=c('init1.txt','init2.txt'))
modelInits(c('init1.txt','init2.txt'))

@@

modelUpdate(12000)
#samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
samplesSet(c("a", "b", "p[]", "psi"))
modelUpdate(10000)
\end{verbatimtab}

@@

eventsToDaysBegin <- c(1, 3, 5, 7, 9, 11)
eventsToDaysEnd   <- c(2, 4, 6, 8, 10, 12)
curDaySinceSnow <- c(1, 2, 1, 2, 1, 2, 1, 2, 1, 2, 1, 2)
daysToObsBegin <- c(1, 4, 7, 10, 13, 16, 19, 22, 25, 28, 31, 34)
daysToObsEnd   <- c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36)
curObserver <- c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3,
                 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3)

# Setting up the parameters.
psi <- .7
beta <- .4
condlOcc <- psi + beta
condlUnocc <- psi

theta <- .5
alpha <- .3
condlTrack <- theta + alpha
condlNoTrack <- theta
p <- c(.70, .80, .90)
E <- c(.05, .02, .03)

@@

# Simulating the data!
for (site in 1:sites) {
    # TRUE == 1, FALSE == 0.
	if (site == 1) {
		occupancy[site] <- 1 * (runif(1) < (condlOcc + condlUnocc/2))
	} else if (occupancy[i-1] == 1) {
		occupancy[site] <- 1 * (runif(1) < condlOcc)
	} else
		occupancy[site] <- 1 * (runif(1) < condlUnocc)
	}
    for (event in 1:snowEvents) {
        otterTrackLaid <- 0
        currentDay <- 0
        # Haha, this is awful.
        for (day in 1:curDaySinceSnow[eventsToDaysBegin[event]]) {
            if (day != 1) {
                currentDay <- currentDay + 1
                otterTrackLaid <- max(otterTrackLaid,
                    occupancy[site] * (runif(1) < theta))
            }
        }
        for (day in eventsToDaysBegin[event]:eventsToDaysEnd[event]) {
            currentDay <- currentDay + 1
            otterTrackLaid <- max(otterTrackLaid,
                occupancy[site] * (runif(1) < theta))
            for (colIndex in daysToObsBegin[day]:daysToObsEnd[day]) {
                numDataPts <- numDataPts + 1
                sawOtterTrack <- otterTrackLaid * 
                    (runif(1) < p[curObserver[colIndex]])
                sawOtherTrack <- (1 - otterTrackLaid) * 
                    (runif(1) < E[curObserver[colIndex]])
                y[numDataPts] <- sawOtterTrack + sawOtherTrack
                siteV[numDataPts] <- site
                eventV[numDataPts] <- event
                dayV[numDataPts] <- currentDay
                obsV[numDataPts] <- curObserver[colIndex]
            }
        }
    }
}
\end{verbatimtab}

\section{Simulated regular data, dependent in $\psi$, $\theta$}
\scriptsize
\begin{verbatimtab}
#Define structure
sites <- 60
events <- 5
snowdays <- 3
obs <- 3
I1 <- rep(c(1,0,0), events*snowdays)
I2 <- rep(c(0,1,0), events*snowdays)

#Define parameters
psi <- .4
beta <- .4
condl.occ <- psi + beta
condl.unocc <- psi
theta <-.3
alpha <-.5
condl.track <- theta + alpha
condl.notrack <- theta
p <- c(.7,.7,.7)
E <- c(.05,.05,.05)

#Initialize variables
occupancy <- NULL
tracks <- matrix(nrow=sites, ncol=events*snowdays)
y <- matrix(nrow=sites, ncol=events*snowdays*obs)

for(i in 1:sites) {
	#Determine occupancy of site. 1*(runif(1)<psi) is 1 around psi*100 percent of the time.
	#Use conditional rates depending on occupancy of site i-1
	if (i==1) 
		{occupancy[i] <- 1*(runif(1)<(condl.occ+condl.unocc)/2)}	
	else if (occupancy[i-1]==1) 
		{occupancy[i] <- 1*(runif(1)<condl.occ)}
	else 
		{occupancy[i] <- 1*(runif(1)<condl.unocc)}
	for(j in 1:events) {
		is.track <- 0
		for(k in 1:snowdays) {
			index <- k + snowdays*(j-1)
			#Decide if a track is laid down on day k of event j, given the occupancy state
			if(i==1) {
				tracks[i,index] <- (runif(1)<(condl.track+condl.notrack)/2)*occupancy[i]
			}else if(tracks[i-1,index]==1) 
				{tracks[i,index] <- (runif(1)<condl.track)*occupancy[i]
			}else 
				{tracks[i,index] <- (runif(1)<condl.notrack)*occupancy[i]}

			#Determine if any tracks have been laid down for event j.
			is.track <- max(is.track, tracks[i,index])
			for(m in 1:obs) {
				index <- m + obs*(k-1) + snowdays*obs*(j-1)
				#Determine if the observer saw a track.
				y[i,index] <- (runif(1)<p[m])*is.track + (runif(1)<E[m])*(1-is.track)
			}
		}
	}
}
\end{verbatimtab}
Note: This needs some more processing before it will run on either model.
\end{document}

>>>>

\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota \\
{\large Senior Comprehensive Exercise: Carleton College}}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

@@

    \begin{figure}
    \begin{center}
        \includegraphics[scale=0.75]{Figures/Diagrams/SimpleHierarchicalModel}
        \caption{A visualization of how the hierarchical model works.}
        \label{simplehierarchicalmodel}
    \end{center}
    \end{figure}

@@

    \begin{figure}
        \centering
        \includegraphics[width=5in]{Figures/R/observerPlots.pdf}
        \caption{The box plots show how the proportion of plots in which an
        observer saw tracks (``occupancy'') differs among the three observers in
        2003. Only those dates on which all three observers flew are included.}
        \label{obsPlots}
    \end{figure}

@@

    for (i in 1:numDataPts) {
        # Positive detection.
        mu[i] <-
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # False detection.
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "HierarchicalModel.txt")
modelCheck("HierarchicalModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal"),
         "OtterData.txt")
modelData("OtterData.txt")
modelCompile(numChains=2)
modelGenInits()

@@

\section{CAR model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    for (site in 1:sites) {
        logit(theta[site]) <- alpha1[site] + alpha0
    }
    for (i in 1:sumNumNeigh) {
        weights[i] <- 1
    }
    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
    alpha0 ~ dflat()
    tau ~ dgamma(.5,.0005)
    psi ~ dunif(0,1)
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    c ~ dgamma(1,.1)
    d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta(a,b) %_% I(.6, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .3)
    }

@@

    for (i in 1:numDataPts) {
        mu[i] <-
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "CARModel.txt")
modelCheck("CARModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal", "num", "adj",
           "sumNumNeigh"),
         "OtterData.txt")
modelData("OtterData.txt")
modelCompile(numChains=2)

init.val <- function() {
list(alpha1 = runif(sites,-100,100),
     alpha0 = runif(1,-100,100),
     tau = rgamma(1,.5,.0005),
     psi = runif(1),
     a = rgamma(1, 1, .1),
     b = rgamma(1, 1, .1),
     c = rgamma(1, 1, .1),
     d = rgamma(1, 1, .1))
}
bugsInits(init.val, numChains=2, fileName=c('init1.txt','init2.txt'))
modelInits(c('init1.txt', 'init2.txt'))

@@

modelUpdate(12000)
samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
modelUpdate(10000)
\end{verbatimtab}

@@

# This is an example of regular data.  Three observers, six snow events, two
# observations on consecutive days after each snow event.
eventsToDaysBegin <- c(1, 3, 5, 7, 9, 11)
eventsToDaysEnd   <- c(2, 4, 6, 8, 10, 12)
curDaySinceSnow <- c(1, 2, 1, 2, 1, 2, 1, 2, 1, 2, 1, 2)
daysToObsBegin <- c(1, 4, 7, 10, 13, 16, 19, 22, 25, 28, 31, 34)
daysToObsEnd   <- c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36)
curObserver <- c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3,
                 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3)

# Setting up the parameters.
psi <- .7
beta <- .4
theta <- .5
p <- c(.70, .80, .90)
E <- c(.05, .02, .03)

@@

# Simulating the data!
for (site in 1:sites) {
    occupancy[site] <- 1 * (runif(1) < psi)
    for (event in 1:snowEvents) {
        otterTrackLaid <- 0
        currentDay <- 0
        for (day in 1:curDaySinceSnow[eventsToDaysBegin[event]]) {
            if (day != 1) {
                currentDay <- currentDay + 1
                otterTrackLaid <- max(otterTrackLaid,
                    occupancy[site] * (runif(1) < theta))
            }
        }
        for (day in eventsToDaysBegin[event]:eventsToDaysEnd[event]) {
            currentDay <- currentDay + 1
            otterTrackLaid <- max(otterTrackLaid,
                occupancy[site] * (runif(1) < theta))
            for (colIndex in daysToObsBegin[day]:daysToObsEnd[day]) {
                numDataPts <- numDataPts + 1
                sawOtterTrack <- otterTrackLaid *
                    (runif(1) < p[curObserver[colIndex]])
                sawOtherTrack <- (1 - otterTrackLaid) *
                    (runif(1) < E[curObserver[colIndex]])
                y[numDataPts] <- sawOtterTrack + sawOtherTrack
                siteV[numDataPts] <- site
                eventV[numDataPts] <- event
                dayV[numDataPts] <- currentDay
                obsV[numDataPts] <- curObserver[colIndex]
            }
        }
    }
}
\end{verbatimtab}

\section{Simulated regular data, dependent in $\psi$, $\theta$}
\scriptsize
\begin{verbatimtab}
# Define structure.
sites <- 60
events <- 5
snowdays <- 3
obs <- 3
I1 <- rep(c(1,0,0), events*snowdays)
I2 <- rep(c(0,1,0), events*snowdays)

# Define parameters.
psi <- .4
beta <- .4
condl.occ <- psi + beta
condl.unocc <- psi
theta <-.3
alpha <-.5
condl.track <- theta + alpha
condl.notrack <- theta
p <- c(.7, .7, .7)
E <- c(.05, .05, .05)

# Initialize variables.
occupancy <- NULL
tracks <- matrix(nrow=sites, ncol=events*snowdays)
y <- matrix(nrow=sites, ncol=events*snowdays*obs)

for (i in 1:sites) {
    # Determine occupancy of site. 1 * (runif(1) < psi) is 1 around psi*100
    # percent of the time. Use conditional rates depending on occupancy of site
    # i-1.
    if (i == 1)
        {occupancy[i] <- 1 * (runif(1) < (condl.occ + condl.unocc) / 2)}
    else if (occupancy[i-1] == 1)
        {occupancy[i] <- 1 * (runif(1) < condl.occ)}
    else
        {occupancy[i] <- 1 * (runif(1) < condl.unocc)}

    for (j in 1:events) {
        is.track <- 0
        for (k in 1:snowdays) {
            index <- k + snowdays * (j - 1)
            # Decide if a track is laid down on day k of event j, given the
            # occupancy state.
            if (i == 1) {
                tracks[i, index] <- (runif(1) <
                                     (condl.track + condl.notrack) / 2) *
                                    occupancy[i]
            } else if (tracks[i-1, index] == 1)
                {tracks[i, index] <- (runif(1) < condl.track) * occupancy[i]
            } else
                {tracks[i, index] <- (runif(1) < condl.notrack) * occupancy[i]}

            # Determine if any tracks have been laid down for event j.
            is.track <- max(is.track, tracks[i,index])
            for (m in 1:obs) {
                index <- m + obs * (k - 1) + snowdays * obs * (j - 1)
                # Determine if the observer saw a track.
                y[i, index] <- (runif(1) < p[m]) * is.track +
                               (runif(1) < E[m]) * (1-is.track)
            }
        }
    }
}
\end{verbatimtab}
Note: This needs some more processing before it will run on either model.
\end{document}
